a and generators

### 2022-04-19 documentos/intro.tex
implemento comentarios de Francois y dejo comentarios a Carlos

Quantum correlations \cite{}, including entanglement~\cite{}, are an important
resource for a wide variate of tasks that include teleportation~\cite{},
quantum computation~\cite{}, and others~\cite{}.
However, this resource is also extremely delicate~\cite{}, specially 
for multi-particle systems~ \cite{},  and important part
of the efforts of the community implementing quantum technologies is 
devoted to tackle this issue from an experimental \cite{} and
theoretical \cite{} point of view. 
The process by which quantum correlations are unintentionally dissipated
is called decoherence~\cite{}
One of the main tools to study the effects of decoherence are 
quantum channels. \cpnote{Ver como pegar con el siguiente bicho. Quizá 
modificar la primera frase del siguiente parrafo o pasar a este.}\par
%
The identification of families of quantum channels among the whole convex set
of quantum channels associated to a Hilbert space, is crucial to understand
quantum noise~\cite{davalosdivisibility,ruskai}, open quantum systems
dynamics~\cite{nielsen_chuang_2011,breuerbook}, and recently even coarse
graining~\cite{Duarte2017,pinedacg}. One of the main difficulties is that,
similar to quantum states, the parameters needed to describe quantum channels
increases overwhelmingly rapid respect to the dimension of the Hilbert space.
For a Hilbert space of dimension $d$, the space of quantum channels is
parametrized with $d^4-d^2$ real parameters which are not fully independent
(the complete positivity condition imposes the regions that physical meaning,
see Ref.~\cite{zimansbook})\cpnote{quiza quitar (QQ), o modificiar 
a decir que el numero de parametros crece exponencialmente con el numero 
de particulas.}. Therefore, even the
qubit case is difficult to
analyze having $12$ parameters corresponding to $6$ angles, $3$ components of
the displacement vector and $3$ describing the compression of the Bloch
sphere~\cite{ruskai}\cpnote{QQ}. Thus, understanding families of channels sharing a common
property usually gives a good insight about the jungle of quantum operations.
For the qubit case there are several studies concerning the unital case, which
non-trivial properties can be described using only $3$ parameters, which in
turn form the well known tetrahedron of Pauli
channels~\cite{Rybar2012,ruskai,davalosdivisibility}\cpnote{QQ}. For general
finite-dimensional channels, several families of channels have been defined.
In~Ref.~\cite{nathanson2007pauli} authors define families of unital channels as
convex combinations of quantum-classical channels, in turn defined with
mutually unbiased basis (MUB), are introduced. These channels have properties
in common with unital qubit channels with positive eigenvalues, and give a
generalization of the Bloch sphere in terms of the generators of the MUBs.
Similarly, a generalization of Pauli channels based on mutually unbiased
measurements is introduced and studied in Ref.~\cite{Siudzinska2020}. Other
studies of channels beyond the qubit can be
found~\cite{pub.1117408526,pub.1060516327,pub.1014660854}.
\cpnote{Me gusta la idea de ir sobre un par de familias, pero no veo para que 
mencionar a las MUBs aca, ni paraque introducir la sigla. Ojala podamos 
discutir que familias mencionamos acá}

>>>>

Quantum correlations \cite{}, including entanglement~\cite{}, are an important
resource for a wide variate of tasks that include teleportation~\cite{},
quantum computation~\cite{}, and others~\cite{}.
However, this resource is also extremely delicate~\cite{}, specially 
for multi-particle systems~ \cite{},  and important part
of the efforts of the community implementing quantum technologies is 
devoted to tackle this issue from an experimental \cite{} and
theoretical \cite{} point of view. 
The process by which quantum correlations are unintentionally dissipated
is called decoherence~\cite{}
One of the main tools to study the effects of decoherence are 
quantum channels. \cpnote{Ver como pegar con el siguiente bicho. Quizá 
modificar la primera frase del siguiente parrafo o pasar a este.}\par
%
The identification of families of quantum channels among the whole convex set
of quantum channels associated to a Hilbert space, is crucial to understand
quantum noise~\cite{davalosdivisibility,ruskai}, open quantum systems
dynamics~\cite{nielsen_chuang_2011,breuerbook}, and recently even coarse
graining~\cite{Duarte2017,pinedacg}. One of the main difficulties is that,
similar to quantum states, the parameters needed to describe quantum channels
increases overwhelmingly rapidly with the dimension of the Hilbert space.
For a Hilbert space of dimension $d$, the space of quantum channels is
parametrized with $d^4-d^2$ real parameters which are not fully independent
(the complete positivity condition imposes the regions that have physical meaning,
see Ref.~\cite{zimansbook})\cpnote{quiza quitar (QQ), o modificiar 
a decir que el numero de parametros crece exponencialmente con el numero 
de particulas.}\ddnote{no crece exponencialmente. La expresión no me parece complicada como para dejarla, sospecho que hay lectores que aprenderan algo interesante de aqui, y espero eso los atrape. Ojo tambien con que la observacion de que son muchos parametros que no son completamente libres, es lo que hace no-trivial el problema matematico que resolvemos aqui.}. Therefore, even the
qubit case is difficult to
analyze having $12$ parameters corresponding to $6$ angles, $3$ components of
the displacement vector and $3$ describing the compression of the Bloch
sphere~\cite{ruskai}\cpnote{QQ}. Thus, describing in detail small families of channels having a given property may provide insight into the jungle of quantum operations.
For the qubit case there are several studies concerning the unital case, for which
non-trivial properties can be described using only $3$ parameters, which in
turn form the well known tetrahedron of Pauli
channels~\cite{Rybar2012,ruskai,davalosdivisibility}\cpnote{QQ}. For general
finite-dimensional channels, several families of channels have been defined.
In~Ref.~\cite{nathanson2007pauli} the authors define families of unital channels as
convex combinations of quantum-classical channels, in turn defined with
mutually unbiased basis (MUB), are introduced. These channels have properties
in common with unital qubit channels with positive eigenvalues, and give a
generalization of the Bloch sphere in terms of the generators of the MUBs.
Similarly, a generalization of Pauli channels based on mutually unbiased
measurements is introduced and studied in Ref.~\cite{Siudzinska2020}. Other
studies of channels beyond the qubit can be
found~\cite{pub.1117408526,pub.1060516327,pub.1014660854}.
\cpnote{Me gusta la idea de ir sobre un par de familias, pero no veo para que 
mencionar a las MUBs aca, ni paraque introducir la sigla. Ojala podamos 
discutir que familias mencionamos acá}\ddnote{las siglas si están de mas, puede que simplemente compactemos los trabajos con MUBs y MUMs con todas las citas genericas donde se estudian generlizacones, sin embargo, ese trabajo de MUBs fue importante para nosotros y al final si identificamos un overlap. Ademas las MUBs son conocidas, y siento que si la gente ve cosas familiaires se va a interesar.}
